### 2015-12-09 doc/thesis/intro.tex

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the evolutionary relationships among a group of
organisms or species. Phylogenies consist of nodes, representing organisms or
species (generally, \defn{taxa}), and edges or branches, which link taxa to
their ancestors and descendants. The \defn{tips} of the phylogeny, that is, the
nodes without any descendants, correspond to \defn{extant}, or observed, taxa.
The \defn{internal nodes} correspond to their extinct common ancestors. The
phylogeny may have a \defn{root}, which is a node with no descendants
distinguished as the most recent common ancestor of all the extant taxa. When
such a root exists, the tree is referred to as being \defn{rooted}; otherwise,
it is \defn{unrooted}. The structural arrangement of nodes and edges in the
tree is referred to as its \defn{topology}. The branches of the phylogeny may
have lengths, which can represent either the genetic distance, or the calendar
time, between ancestors and their descendants. Phylogenies whose branch lengths
are in units of calendar time are referred to as \defn{time-scaled}.

In the case of a rooted tree with branch lengths in calendar time, the internal
nodes can be mapped onto a timeline by using the tips of the phylogeny, which
usually map to the present day, as a reference point. These \defn{branching
times} correspond to the times at which speciation events
occurred~\autocite{nee1992tempo}. In phylogenies relating animal species,
speciation most often occurs by an \defn{allopatric} process, where a
sub-population of organisms is geographically isolated from the original
population [a reference to Coyne \& Orr will go here once I skim the relevant
bits of the book]. Over time, the two populations diverge genetically,
eventually resulting in two distinct species.

Many RNA viruses, such as the human immuno-deficiency virus (HIV), evolve
rapidy enough that populations can accumulate measurable genetic changes on
time scales of months or years~\autocite{drummond2003measurably}. Transmission
of the virus from one host to another closely mirrors the allopatric speciation
process. Upon transmission, a sub-population of virions from the donor host is
geographically isolated in the recpient, followed by genetic divergence of the
two hosts' infections. Therefore, the branching time relating the two hosts'
viral populations in a time-scaled phylogeny corresponds to the time of
transmission. This correspondence can be quite rough, and is subject to some
important limitations which are discussed below. However, even taking these
limitations into account, the relationship between branching times in the
phylogeny and transmission times in the population theoretically implies that
the evolutionary history of the virus is informative about its epidemiology. In
other words, information about transmission patterns can be inferred from the
viral phylogeny, complementing more traditional survey-based epidemiological
methods. The tools and techniques for making such inferences are referred to as
\defn{phylodynamic} methods.

>>>>

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the evolutionary relationships among a group of
organisms or species. Phylogenies consist of nodes, representing organisms or
species (generally, \defn{taxa}), and edges or branches, which link taxa to
their ancestors and descendants. The \defn{tips} of the phylogeny, that is, the
nodes without any descendants, correspond to \defn{extant}, or observed, taxa.
The \defn{internal nodes} correspond to their extinct common ancestors. The
phylogeny may have a \defn{root}, which is a node with no descendants
distinguished as the most recent common ancestor of all the extant taxa. When
such a root exists, the tree is referred to as being \defn{rooted}; otherwise,
it is \defn{unrooted}. The structural arrangement of nodes and edges in the
tree is referred to as its \defn{topology}. The branches of the phylogeny may
have lengths, which can represent either the genetic distance, or the calendar
time, between ancestors and their descendants. Phylogenies whose branch lengths
are in units of calendar time are referred to as \defn{time-scaled}. In the
case of a rooted tree with branch lengths in calendar time, the internal nodes
can be mapped onto a timeline by using the tips of the phylogeny, which usually
map to the present day, as a reference point. These \defn{branching times}
correspond to the times at which speciation events
occurred~\autocite{nee1992tempo}. These concepts are illustrated in
Figure~\ref{fig:phylogeny}.

\begin{figure}
  \label{fig:phylogeny}
  \includegraphics{phylogeny}
  \caption{Cartoon illustration of a rooted, time-scaled phylogeny relating
  several domestic animals. The tips of the phylogeny, which represent extant
  taxa, are placed at the present day on the time axis. Internal nodes,
  representing extinct common ancestors to the extant taxa, fall in the past,
  at branching times corresponding to speciation events. The topology of the
  tree indicates that cats and dogs are the most closely related pair of
  species, whereas fish is most distantly related to any other node in the
  tree.}
\end{figure}

Various population genetic processes cause the phylogenies that we observe to
take the particular shapes that they do. Selection, due to environmental
pressures, allows certain lineages to persist while driving others to
extinction. Speciation, which causes branching, often occurs by an
\defn{allopatric} process, where a sub-population of organisms is isolated from
the original population by a geographic barrier~\autocite{coyne2004speciation}.
Over time, the two populations diverge genetically, eventually resulting in two
distinct species. Barriers are not necessary for speciation, however.
Diversifying selection, which favours either of two ``extreme'' phenotypes, can
cause \defn{sympatric} speciation without a geographic
separation~\autocite{coyne2004speciation}.

Now, consider a phylogeny not of macro-organisms, but of RNA viruses. Many of
these, such as the human immuno-deficiency virus (HIV), evolve rapidy enough
that populations can accumulate measurable genetic changes on time scales of
months or years~\autocite{drummond2003measurably}. Because viruses cannot exist
for long outside their hosts, the host replaces the outside world as the
``environment'' in which evolution takes place. In this light, striking
parallels can be drawn between immunological and epidemiological processes at
the host level, and the evolutionary processes studied in population
genetics~\autocite{grenfell2004unifying}. For example, both the host immune
response~\autocite{??} and the presence of anti-retroviral drugs~\autocite{??}
exert directional selection on the viral population within a single host.
Transmission, where a sub-population of virions from the donor host is
``geographically'' isolated in the recpient, closely mirrors the allopatric
speciation process and causes a branching point in the viral phylogeny
(although this may not correspond exactly to the time of transmission, as
discussed in the next section). Consequently, any factor which influences
transmission dynamics, such as the virus' mode of transmission or the
population structure of its hosts, in turn affects the shape of the phylogeny.
The study of this interaction between host-level factors and viral phylogies is
known as \defn{phylodynamics}~\autocite{grenfell2004unifying}.

In particular, if a time-scaled phylogeny is constructed where all viruses from
the same host are collapsed into a single taxon, we might expect the branching
time relating the two hosts to correspond to the time of transmission. Of
course, this will only be the case if the two hosts' infections indeed diverged
allopatrically at the time of transmission, and not sympatrically within the
donor host at some previous time. This limitation, and others, in interpreting
the branching times of the viral phylogeny as transmission times, are discussed
in the next section. Nevertheless, even taking these limitations into account,
the relationship between branching times in the phylogeny and transmission
times in the population theoretically implies that the evolutionary history of
the virus is informative about the biology and epidemiology of its hosts. In
other words, information about, for example, transmission patterns, might be
inferred from the viral phylogeny. Phylodynamic methods are intended to
complement more traditional approaches to gathering epidemiological data, such
as population surveys.
